Airspace configuration decision based on Monte-Carlo tree search

### 2018-04-23 article/article.tex
(~) Refactored notations
No action, only nodes

\section*{Notations}
\begin{compactitem}
  \item $m$: elementary module;
  \item $S$: sector control position, group of modules;
  \item $P$: partition of the airspace, group of sectors;
  \item $t$: current time;
  \item $C(P, t)$: cost of a given partition $P$ at time $t$;
  \item \(\mathcal{S}\): the set of available states;
  \item \(\mathcal{A}\): the set of all possible actions;
  \item \(\mathcal{A}(s)\): the available actions from a state \(s\);
  \item $s$: state;
  \item \(f \colon \mathcal{S} \times \mathcal{A} \to \mathcal{S}\):
    transition function giving the resulting states of action \(a\) executed
    from a state \(s\).
  \item $h \colon \mathcal{S} \to \mathbb{R}^+$: heuristic estimating the cost
    from a state \(s \in \mathcal{S}\) to a final state;
  \item \(\mathcal{N}\): the set of nodes
  \item \(u, v \in \mathcal{N}^2\): nodes
  \item \(Q \colon \mathcal{N} \to \mathbb{R}^+\): the reward expectancy of a
    node;
  \item \(N \colon \mathcal{N} \to \mathbb{N}\): number of simulations running
    through or from a node;
  \item \(\first(\ell)\): returns the first element of \(\ell\).
\end{compactitem}

@@

The above mentioned steps are denoted
\begin{compactenum}
\item selection (or tree walk): choosing successively most promising nodes,
\item expansion: adding new nodes to the search tree,
\item simulation (or random walk): choosing quickly actions from the model
  tree,
\item backpropagation (or backup): applying the result of the simulation to
  the search tree.
\end{compactenum}
Those phases are repeated until a stopping criterium (e.g.\ memory or time) is
reached, resulting in algorithm~\ref{alg:gen_mcts} where the tree policy
aggregates phases 1 and 2 to create a new node of the search tree and the
default policy gives an evaluation of the newly added node.
\begin{algorithm}
  \caption{General MCTS~\cite{browne2012survey}}\label{alg:gen_mcts}
  \begin{algorithmic}
    \Procedure{MctsSearchTree}{$v_0$}
    \While{within computational budget}
    \State{} \([v_0, \dots, v_n] \gets\) \Call{TreePolicy}{$v_0$}
    \State{} \(\Delta \gets\) \Call{DefaultPolicy}{$v_n$}
    \State{} \Call{Backup}{$[v_0, \dots, v_n]$, $\Delta$}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsubsection{Extending the search tree}
\paragraph{Selection (algorithm~\ref{alg:tree_pol})}
The aim of the selection is to choose from a set of actions, the most promising
one\footnote{i.e.\ possibly leading to the best evaluation} according to a
\emph{tree policy}. This type of problem can be solved by bandits methods. Those
methods consist in, given a bandit in front of several slot machines (multi
armed bandit), decide which machine will bring the highest reward knowing the
past result. The objective of bandits methods is thus to maximise the reward and
minimise the regret of not playing the best machine.

The bandit problem has been applied to MCTS via the Upper Confidence Tree
algorithm in~\cite{kocsis2006bandit} using the UCB1 equation~\ref{eq:ucb1}.
Let \(v\) be the node from which an action must be
selected to go deeper in the tree, \(a\) an action available from \(v\),
\(\mu_{v,a}\) the expected outcome from the node obtained by applying action
\(a\) on \(v\), \(T_v\) the number of simulations carried out from node \(v\)
(which includes any simulation from nodes in any subtree of \(v\)) and
\(T_{v,a}\) the number of simulations carried out from the node resulting from
the action \(a\) applied to node \(v\), \(\beta\) a constant. The selected node
is thus the one maximising the \emph{UCB1} equation
\begin{equation}
  \label{eq:ucb1}
  \mu_{v,a} + 2 \beta \sqrt{\frac{2 \log T_v}{T_{v,a}}}
\end{equation}

While the previous equation is well suited for two players games, it can be
tweaked to improve its efficiency in one player games or sequencing problems. An
alternative using the standard deviation of the outcome of the simulations is
proposed in~\cite{sebag2010fuse} called the \emph{UCB1-tuned} equation
\begin{equation}
  \mu_{v,a} +
  \sqrt{%
    \frac{\beta\log T_v}{T_{v,a}}
    \min\left( \frac{1}{4}, \sigma^2_{v,a} +
    \sqrt{%
      \frac{2 \log T_v}{T_{v,a}}
    }\right)
  }
\end{equation}
\begin{algorithm}
  \caption{UCT algorithm}\label{alg:tree_pol}
  \begin{algorithmic}
    \Function{TreePolicy}{$\pi$}
    \State{}\(v \gets \first(\pi)\)
    \If{$v$ is terminal}
    \State{}\Return{$\pi$}
    \Else{}
    \If{$v$ is not fully expanded}
    \State{}\Return{\Call{Expand}{$v$} $\cup \pi$}
    \Else{}
    \State{}$f \gets$ \Call{BestChild}{$v$}
    \State{}\Return{%
      \Call{TreePolicy}{$f \cup \pi$}
    }
    \EndIf{}
    \EndIf{}
    \EndFunction{}
  \end{algorithmic}
  \begin{algorithmic}
    \Function{BestChild}{$v$}
    \State{}\Return{$\argmax\{\Call{UCB}{v'} | v' \text{\,children of\,} v \}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

\paragraph{Expansion (algorithm~\ref{alg:expansion})}
The above step is carried out only if, considering a node \(v\), a simulation
has been run through each child available from \(v\). In the case where one or
more children haven't been explored yet, one of the unexplored children is
chosen and a simulation is started from it.
\begin{algorithm}
  \caption{Expansion}\label{alg:expansion}
  \begin{algorithmic}
    \Function{Expand}{$v$}
    \State{}$\ell \gets \{ v | v \text{\,children of\,} u, N(v) = 0 \}$
    \State{}\Return{random element from $\ell$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

@@

\paragraph{Backpropagation}
The backpropagation consists in updating the values required to carry out the
tree policy. The values must be updated incrementally since the backpropagation
function has the current value of the parameters and the result of the
simulation as parameters. Thus, for UCB1 equation, the expected reward (mean)
and the total count are updated this way
\begin{algorithm}
  \caption{UCB1 backpropagation}
  \begin{algorithmic}
    \Procedure{Backpropagate}{$v$, $r$}
    \State{}\(\delta \gets\ r - \mu_v\)
    \State{}\(Q(v) \gets Q(v) + \frac{\delta}{N(v) + 1}\)
    \State{}\(N(v) \gets N(v) + 1\)
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

For the UCB1-tuned, the standard deviation has to be computed. It results in
algorithm~\ref{alg:backpropagate-tuned} which introduces the value \(m_2(v)\)
associated to a node \(v\) to compute the standard deviation via the formula
\begin{equation}
  \sigma^2_v = \frac{m_2(v)}{N(v)}.
\end{equation}
\begin{algorithm}
  \caption{UCB1-tuned backpropagation}\label{alg:backpropagate-tuned}
  \begin{algorithmic}
    \Procedure{Backpropagate}{$v, r$}
    \State{}\(\delta \gets r - Q(v)\)
    \State{}\(Q(v) \gets Q(v) + \frac{\delta}{N(v) + 1}\)
    \State{}\(\delta' \gets r - Q(v)\)
    \State{}\(m_2(v) \gets m_2(v) + \delta \delta'\)
    \State{}\(N(v) \gets N(v) + 1\)
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsection{Sequence building}
\subsubsection{Node selection}\label{sssec:node_selection}
The building of the final sequence of actions differs from the selection on the
aim, the selection chooses the most promising actions while the final sequence
must be composed of the best. This means that the final path build won't use the
UCB equation. Chaslot \textit{et al.}~\cite{browne2012survey} propose several
methods to select nodes,
\begin{compactitem}
  \item max-child: select the child with highest reward;
  \item robust child: select the most visited root child;
  \item secure child: select the child maximising a lower confidence bound.
\end{compactitem}

\subsubsection{Iterative pathfinding}
The path is built iteratively by calling successive Monte Carlo tree searches.
Say an MCTS has been called on a node \(u\). Once a stopping criterium is
matched, a node \(v\) among the ones reachable from \(u\) is selected via one of
the previously mentioned policies. Then an mcts is called back with node \(u\)
as the root node. The path is composed of the successive roots. The algorithm is
described at~\ref{alg:path_building}.
\begin{algorithm}
  \caption{%
    Path building. The {\sc NextNode} (here max-child) function is one among
    those in~\ref{sssec:node_selection}.
  }\label{alg:path_building}
  \begin{algorithmic}
    \Function{MctsSearch}{$u_0, n$}
    \State{}\(u \gets u_0\)
    \State{}\(\pi \gets \{u\}\)
    \For{$i = 1$ to $n$}
      \State{}\Call{MctsSearchTree}{$u$}
      \State{}\(u \gets\)\Call{NextNode}{$u$}
      \State{}\(\pi \gets \pi \cup \{u\}\)
    \EndFor{}
    \State{}\Return{$\pi$}
    \EndFunction{}
    \Function{NextNode}{$u$}
    \State{}\Return{$\argmax\{Q(v) | v \text{\,children of\,} u\}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}
\subsection{A\(^*\)}
To evaluate the exactness of the paths given by the MCTS algoithm, the A\(^*\)
algorithm is used. A\(^*\) is an exact depth first search pathfinding algorithm
which uses a heuristic function to guide its search. The algorithm is given
in~\ref{alg:astar} where \(u_0\) is the initial state, \(T\) the set of termial
nodes, \(h\) a heuristic function estimating the cost from a state \(u\) given
as argument to a final state, \(k \colon \mathcal{S}^2 \to \mathbb{R}\). The
function \(\mathup{first(G)}\) returns the first element of \(G\) and
\(f\mathup{-insert}(G, v)\) inserts \(v\) in \(G\) ordering first by \(f\)
increasing then by \(g\) decreasing.
\begin{algorithm}
  \caption{A\(^*\) algorithm~\cite{alliotschiex2002ia&it}}\label{alg:astar}
  \begin{algorithmic}
    \Procedure{A$^*$}{$u_0$}
    \State{}\(G \gets u_0; D \gets \emptyset; g(u_0) \gets 0; f(u_0) \gets 0\)
    \State{}\(\mathup{father}(u_0) \gets \emptyset\)
    \While{$G \neq \emptyset$}
    \State{}\(u \gets \mathup{first}(G); G \gets G \backslash \{u\}\)
    \State{}\(D \gets D \cup \{u\}\)
    \If{$u \in T$}
    \State{}\Return{father}
    \EndIf{}
    \For{$v$ in childen of $u$}
    \If{$v \notin D \cup G$ or $[g(v) > g(u) + k(u, v)]$}
    \State{}\(g(v) \gets g(u) + k(u, v)\)
    \State{}\(f(v) \gets g(v) + h(v)\)
    \State{}\(\mathup{father}(v) \gets u\)
    \State{}\(f\mathup{-insert}(G, v)\)
    \EndIf{}
    \EndFor{}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}
\section{Model}

>>>>

\section*{Notations}
\begin{compactitem}
  \item $m$: elementary module;
  \item $S$: sector control position, group of modules;
  \item $P$: partition of the airspace, group of sectors;
  \item $t$: current time;
  \item $C(P, t)$: cost of a given partition $P$ at time $t$;
  \item \(u, v \in \mathcal{N}^2\): nodes
  \item \(\mathcal{N}\): the set of nodes;
  \item $h \colon \mathcal{N} \to \mathbb{R}^+$: heuristic estimating the cost
    from a node \(u\) to a final node;
  \item \(\mu_v\): mean value of outcomes of simulations run through or from a
    node \(v\);
  \item \(T_v\): number of simulations run through or from a node \(v\);
  \item \(\first(\ell)\): returns the first element of \(\ell\).
\end{compactitem}

@@

The above mentioned steps are denoted
\begin{compactenum}
\item selection (or tree walk): choosing successively most promising nodes,
\item expansion: adding new nodes to the search tree,
\item simulation (or random walk): choosing successively nodes from the model
  tree,
\item backpropagation (or backup): applying the result of the simulation to
  the search tree.
\end{compactenum}
Those phases are repeated until a stopping criterium (e.g.\ memory or time) is
reached, resulting in algorithm~\ref{alg:gen_mcts} where the tree policy
aggregates phases 1 and 2 to create a new node of the search tree and the
default policy gives an evaluation of the newly added node.
\begin{algorithm}
  \caption{General MCTS~\cite{browne2012survey}}\label{alg:gen_mcts}
  \begin{algorithmic}
    \Procedure{MctsSearchTree}{$v_0$}
    \While{within computational budget}
    \State{} \([v_0, \dots, v_n] \gets\) \Call{TreePolicy}{$v_0$}
    \State{} \(\Delta \gets\) \Call{DefaultPolicy}{$v_n$}
    \State{} \Call{Backup}{$[v_0, \dots, v_n]$, $\Delta$}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsubsection{Extending the search tree}
\paragraph{Selection (algorithm~\ref{alg:tree_pol})}
The aim of the selection is to choose the most promising node among the children
available\footnote{i.e.\ possibly leading to the best evaluation} according to
a \emph{tree policy}. This type of problem can be solved by bandits methods.
Those methods consist in, given a bandit in front of several slot machines
(multi armed bandit), decide which machine will bring the highest reward knowing
the past result. The objective of bandits methods is thus to maximise the reward
and minimise the regret of not playing the best machine.

The bandit problem has been applied to MCTS via the Upper Confidence Tree
algorithm in~\cite{kocsis2006bandit} using the UCB1 equation~\ref{eq:ucb1}.
Let \(u\) be the node from which a child \(v\) must be selected to go deeper in
the tree, \(T_u\) the number of simulations carried out from node \(u\)
(which includes any simulation from nodes in any subtree of \(u\)) and
\(\beta\) a constant. The selected child \(v\) is thus the one maximising the
\emph{UCB1} equation
\begin{equation}
  \label{eq:ucb1}
  \mu_v + 2 \beta \sqrt{\frac{2 \log T_u}{T_v}}
\end{equation}

While the previous equation is well suited for two players games, it can be
tweaked to improve its efficiency in one player games or sequencing problems. An
alternative using the standard deviation of the outcome of the simulations is
proposed in~\cite{sebag2010fuse} called the \emph{UCB1-tuned} equation
\begin{equation}
  \mu_v +
  \sqrt{%
    \frac{\beta\log T_u}{T_v}
    \min\left( \frac{1}{4}, \sigma^2_v +
    \sqrt{%
      \frac{2 \log T_u}{T_v}
    }\right)
  }
\end{equation}
\begin{algorithm}
  \caption{UCT algorithm}\label{alg:tree_pol}
  \begin{algorithmic}
    \Function{TreePolicy}{$\pi$}
    \State{}\(v \gets \first(\pi)\)
    \If{$v$ is terminal}
    \State{}\Return{$\pi$}
    \Else{}
    \If{$v$ is not fully expanded}
    \State{}\Return{\Call{Expand}{$v$} $\cup \pi$}
    \Else{}
    \State{}$f \gets$ \Call{BestChild}{$v$}
    \State{}\Return{%
      \Call{TreePolicy}{$f \cup \pi$}
    }
    \EndIf{}
    \EndIf{}
    \EndFunction{}
  \end{algorithmic}
  \begin{algorithmic}
    \Function{BestChild}{$v$}
    \State{}\Return{$\argmax\{\Call{UCB}{v'} | v' \text{\,children of\,} v \}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

\paragraph{Expansion (algorithm~\ref{alg:expansion})}
The above step is carried out only if, considering a node \(v\), a simulation
has been run through each child available from \(v\). In the case where one or
more children haven't been explored yet, one of the unexplored children is
chosen and a simulation is started from it.
\begin{algorithm}
  \caption{Expansion}\label{alg:expansion}
  \begin{algorithmic}
    \Function{Expand}{$v$}
    \State{}$\ell \gets \{ v | v \text{\,children of\,} u, T_v = 0 \}$
    \State{}\Return{random element from $\ell$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

@@

\paragraph{Backpropagation}
The backpropagation consists in updating the values required to carry out the
tree policy. The values must be updated incrementally since the backpropagation
function has the current value of the parameters and the result of the
simulation as parameters. Thus, for UCB1 equation, the expected reward (mean)
and the total count are updated this way
\begin{algorithm}
  \caption{UCB1 backpropagation}
  \begin{algorithmic}
    \Procedure{Backpropagate}{$u$, $r$}
    \State{}\(\delta \gets r - \mu_u\)
    \State{}\(\mu_u \gets \mu_u + \frac{\delta}{T_u + 1}\)
    \State{}\(T_u \gets T_u + 1\)
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

For the UCB1-tuned, the standard deviation has to be computed. It results in
algorithm~\ref{alg:backpropagate-tuned} which introduces the value \(m_2(u)\)
associated to a node \(u\) to compute the standard deviation via the formula
\begin{equation}
  \sigma^2_u = \frac{m_2(u)}{T_u}.
\end{equation}
\begin{algorithm}
  \caption{UCB1-tuned backpropagation}\label{alg:backpropagate-tuned}
  \begin{algorithmic}
    \Procedure{Backpropagate}{$u, r$}
    \State{}\(\delta \gets r - \mu_u\)
    \State{}\(\mu_u \gets \mu_u + \frac{\delta}{T_u + 1}\)
    \State{}\(\delta' \gets r - \mu_u\)
    \State{}\(m_2(u) \gets m_2(u) + \delta \delta'\)
    \State{}\(T_u \gets T_u + 1\)
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsection{Sequence building}
\subsubsection{Node selection}\label{sssec:node_selection}
The building of the final sequence of nodes differs from the selection on the
aim, the selection chooses the most promising node while the final sequence
must be composed of the best. This means that the final path build won't use the
UCB equation. Chaslot \textit{et al.}~\cite{browne2012survey} propose several
methods to select nodes,
\begin{compactitem}
  \item max-child: select the child with highest reward;
  \item robust child: select the most visited root child;
  \item secure child: select the child maximising a lower confidence bound.
\end{compactitem}

\subsubsection{Iterative pathfinding}
The path is built iteratively by calling successive Monte Carlo tree searches.
Say an MCTS has been called on a node \(u\). Once a stopping criterium is
matched, a node \(v\) among the ones reachable from \(u\) is selected via one of
the previously mentioned policies. Then an mcts is called back with node \(u\)
as the root node. The path is composed of the successive roots. The algorithm is
described at~\ref{alg:path_building}.
\begin{algorithm}
  \caption{%
    Path building. The {\sc NextNode} (here max-child) function is one among
    those in~\ref{sssec:node_selection}.
  }\label{alg:path_building}
  \begin{algorithmic}
    \Function{MctsSearch}{$u_0, n$}
    \State{}\(u \gets u_0\)
    \State{}\(\pi \gets \{u\}\)
    \For{$i = 1$ to $n$}
      \State{}\Call{MctsSearchTree}{$u$}
      \State{}\(u \gets\)\Call{NextNode}{$u$}
      \State{}\(\pi \gets \pi \cup \{u\}\)
    \EndFor{}
    \State{}\Return{$\pi$}
    \EndFunction{}
    \Function{NextNode}{$u$}
    \State{}\Return{$\argmax\{\mu_v | v \text{\,children of\,} u\}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}
\subsection{A\(^*\)}
To evaluate the exactness of the paths given by the MCTS algoithm, the A\(^*\)
algorithm is used. A\(^*\) is an exact depth first search pathfinding algorithm
which uses a heuristic function to guide its search. The algorithm is given
in~\ref{alg:astar} where \(u_0\) is the initial state, \(T\) the set of termial
nodes, \(h\) a heuristic function estimating the cost from a state \(u\) given
as argument to a final state, \(k \colon \mathcal{S}^2 \to \mathbb{R}\). The
function \(\mathup{first(G)}\) returns the first element of \(G\) and
\(f\mathup{-insert}(G, v)\) inserts \(v\) in \(G\) ordering first by \(f\)
increasing then by \(g\) decreasing.
\begin{algorithm}
  \caption{A\(^*\) algorithm~\cite{alliotschiex2002ia&it}}\label{alg:astar}
  \begin{algorithmic}
    \Procedure{A$^*$}{$u_0$}
    \State{}\(G \gets u_0; D \gets \emptyset; g(u_0) \gets 0; f(u_0) \gets 0\)
    \State{}\(\mathup{father}(u_0) \gets \emptyset\)
    \While{$G \neq \emptyset$}
    \State{}\(u \gets \mathup{first}(G); G \gets G \backslash \{u\}\)
    \State{}\(D \gets D \cup \{u\}\)
    \If{$u \in T$}
    \State{}\Return{father}
    \EndIf{}
    \For{$v$ in childen of $u$}
    \If{$v \notin D \cup G$ or $[g(v) > g(u) + k(u, v)]$}
    \State{}\(g(v) \gets g(u) + k(u, v)\)
    \State{}\(f(v) \gets g(v) + h(v)\)
    \State{}\(\mathup{father}(v) \gets u\)
    \State{}\(f\mathup{-insert}(G, v)\)
    \EndIf{}
    \EndFor{}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}
\section{Model}
